Spectroscopic confirmation of very-wide stellar binaries and large-separation comoving pairs from \textsl{Gaia DR1}

### 2017-10-10 paper/comoving-followup.tex
wrapping
referee minor concerns

\begin{abstract}
% Context
Members of widely-separated stellar binaries ($\gtrsim 0.1$~pc) or loosely bound
stellar associations and clusters are born with small spreads in kinematics and
chemistry, but are thought to physically dissociate over short timescales.
The kinematic properties of comoving stars---bound, wide or unbound but coeval
systems---are therefore sensitive to both the smooth tidal field and the
spectrum of massive perturbers (e.g., stellar remnants, dark matter, giant
molecular clouds) in the Milky Way.
% Aims
% Thus far, a few hundred confirmed comoving pairs with separations as large as
% $8~\pc$ have been discovered in the solar neighborhood using
% \project{Hipparcos} astrometry combined with radial velocity surveys.
In previous work, we identified $>4,000$ candidate comoving star pairs using
only astrometric data from the Tycho-Gaia Astrometric Solution (TGAS) catalog
and found a large number of candidate comoving pairs with separations $\gtrsim
1~\pc$.
Without radial velocity measurements for the vast majority of these pairs, the
false-positive rate at large separations is significant ($\sim 60\%$).
% Methods
Here we present results from our own low-resolution radial-velocity survey of a
random sample of the candidate pairs to identify high-probability comoving star
pairs.
% Results
Of the \npairsobs\ observed comoving pairs, \nhighprob\ comoving pairs have
probabilities $50\%$ with separations as large as $\approx 10~\pc$, our original
search limit.
Between separations of $10^{-3}~\pc$ to $\approx 0.5~\pc$,
the number of high-probability pairs per separation decreases with increasing
separation.
From $\approx 0.5~\pc$ to $10~\pc$, the number per separation is approximately
uniform.
% Conclusions
We thus confirm the discovery of a population of comoving star pairs at very
large separations, suggesting that disrupted wide binaries and stellar
associations remain approximately spatially coherent.
\todo{something about stellar models and chemical tagging}
\end{abstract}

@@

With large samples of candidate pairs, several groups have measured the
separation distribution of wide binaries at angular separations $\Delta \theta
\lesssim 1^\circ$ where the false-positive rate is expected to be lower
and the samples are likely predominantly wide binaries
(\citealt{Chaname:2004,Lepine:2007,Sesar:2008}).
In most cases, the angular or projected separation distribution is fit using a
decreasing power-law, and the power-law index, $\alpha$, is found to be $\alpha
\approx 1.5$--$2$ out to angular separations of $\approx 0.25^\circ$ ($\approx
0.65~\pc$ at a distance of $150~\pc$), apparently consistent with simulations
(e.g., \citealt{Weinberg:1987}).
Tentative reports of a steepening or cutoff of the separation distribution at
larger separations have been claimed (e.g., \citealt{Yoo:2004,Quinn:2009}), but
these claims are limited by small numbers of high-probability pairs at large
separations.

@@

Indeed, observational work that followed appears to find tentative evidence for
the existence of a second peak in the separation distribution of
radial-velocity-confirmed comoving stars (\citealt{Shaya:2011}; hereafter SO11).
A key aspect of this work was to move from searching for common proper motion
pairs to performing a statistical inference of the difference in full-space
velocity vector for pairs of stars.
At separations of $1$--$10~\pc$, star pairs in the local solar neighborhood
subtend large angles on the sky: even if the two component stars were moving
with the same 3D velocity, their proper motions and radial velocities can be
different.
By compiling radial velocities from the literature for \project{Hipparcos}
stars, SO11 applied this method to search for all comoving star pairs with
separations larger than $0.01~\pc$ within $100~\pc$ of the sun.
They find $\sim 200$ high probability comoving pairs with a separation
distribution that falls of until $\sim 1~\pc$, where the number of pairs then
begins rising in bins of log-separation.

@@

To select candidate pairs for follow-up spectroscopic observations, we only
consider mutually exclusive star pairs within $200~\pc$ in Heliocentric
distance.
With the SNR cut used above and given the limitations of the \tgas\ catalog
itself, \tgas\ is $\approx$90\% complete to $200~\pc$ (\citealt{Bovy:2017}).
From this source target list, we randomly observed targets with airmass $\sec z
< 1.5$ during our observing run.
\figurename~\ref{fig:sample-cmd} shows a color-magnitude diagram of all targets
observed in this observing run (markers) plotted over the density of 195,294
\tgas\ stars within $200~\pc$ with \acronym{2MASS} photometry.
Candidate comoving pairs are connected by (arbitrarily colored) lines.

@@

Spectra were obtained using the \project{Modspec} spectrograph mounted on the
$2.4~{\rm m}$ Hiltner telescope at MDM
observatory\footnote{\url{http://mdm.kpno.noao.edu}} on Kitt Peak (Arizona).
\project{Modspec} was set up in long-slit mode with a 1\arcsec\ slit, a
$600~{\rm line}~{\rm mm}^{-1}$ grating, and a $2048\times2048~{\rm pixel}$ SITe
CCD detector (``Echelle'').
Only the central 300 pixel columns were read out from the CCD, and pixel rows
$>1600$ (at the blue end) were later trimmed because of the poor quantum
efficiency of the CCD at wavelengths $\lesssim 3600~{\rm \AA}$.
The resolution and wavelength range were $\approx2~{\rm \AA}~{\rm pixel}^{-1}$
from $\approx 3600$--$7200~{\rm \AA}$.
At this resolution, to obtain high signal-to-noise ($>100$ per pixel) spectra,
most exposures were between $30$--$120~{\rm s}$.
To maximize the number of observed pairs, we chose not to take
comparison lamp spectra at each pointing as it adds an overhead of $\approx
90~{\rm s}$ per observation.
We are primarily interested in relative velocities between pair members and
therefore don't need precise absolute calibrations of their radial velocities.
To determine rough nightly wavelength solutions, calibration spectra were
obtained using Hg-Ne and Ne comparison lamps at the beginning and end of each
night.
Additionally, atmospheric emission lines were used to correct the wavelength
solution on a per-object basis.

@@

\begin{table}[h]
    \centering
    \caption{Description of columns in data table containing separations and
    relative radial velocities for all observed groups, along with likelihood
    ratios computed with and without relative radial velocities.}
    \label{tbl:data-pairs}
    \begin{tabular}{l|l|l}
        \toprule
        Column Name    & Unit & Description\\
        \midrule
        \texttt{Oh17\_group\_id}    &      & Group ID from candidate catalog (\citealt{Oh:2017})\\
        \texttt{prob}               &      & \todo{Probability ...}\\
        \texttt{sep\_tan}           & \pc  & Median of tangential separation error distribution\\
        \texttt{sep\_tan\_err}      & \pc  & Uncertainty in tangential separation\\
        \texttt{relative\_rv}       & \kms & Relative radial velocity between the two stars\\
        \texttt{relative\_rv\_err}  & \kms & Uncertainty in relative radial velocity\\
        \bottomrule
    \end{tabular}
\end{table}

>>>>

\begin{abstract}
% Context
Members of widely-separated stellar binaries ($\gtrsim 0.1$~pc) or loosely bound
stellar associations and clusters are born with small spreads in kinematics and
chemistry, but are thought to physically dissociate over short timescales.
The kinematic properties of comoving stars---bound, wide or unbound but coeval
systems---are therefore sensitive to both the smooth tidal field and the
spectrum of massive perturbers (e.g., stellar remnants, dark matter, giant
molecular clouds) in the Milky Way.
% Aims
% Thus far, a few hundred confirmed comoving pairs with separations as large as
% $8~\pc$ have been discovered in the solar neighborhood using
% \project{Hipparcos} astrometry combined with radial velocity surveys.
In previous work, we identified $>4,000$ candidate comoving star pairs using
only astrometric data from the Tycho-Gaia Astrometric Solution (TGAS) catalog
and found a large number of candidate comoving pairs with separations $\gtrsim
1~\pc$.
Without radial velocity measurements for the vast majority of these pairs, the
false-positive rate at large separations is significant ($\sim 60\%$).
% Methods
Here we present results from our own low-resolution radial-velocity survey of a
random sample of the candidate pairs to identify high-probability comoving star
pairs.
% Results
Of the \npairsobs\ observed comoving pairs, \nhighprob\ comoving pairs have
probabilities $50\%$ with separations as large as $\approx 10~\pc$, our original
search limit.
Between separations of $10^{-3}~\pc$ to $\approx 0.5~\pc$,
the number of high-probability pairs per separation decreases with increasing
separation.
From $\approx 0.5~\pc$ to $10~\pc$, the number per separation is approximately
uniform.
% Conclusions
We thus confirm the discovery of a population of comoving star pairs at very
large separations, suggesting that disrupted wide binaries and stellar
associations remain approximately spatially coherent.
% \todo{something about stellar models and chemical tagging?}
\end{abstract}

@@

With large samples of candidate pairs, several groups have measured the
separation distribution of wide binaries at angular separations $\Delta \theta
\lesssim 1^\circ$ where the false-positive rate is expected to be lower
and the samples are likely predominantly wide binaries
(\citealt{Chaname:2004,Lepine:2007,Sesar:2008}).
In most cases, the angular or projected separation distribution is fit using a
decreasing power-law, and the power-law index, $\alpha$, is found to be $\alpha
\approx 1.5$--$2$ out to separations $\approx 0.65~\pc$ (angular separations of
$\approx 0.25^\circ$ at a distance of $150~\pc$), apparently consistent with
simulations (e.g., \citealt{Weinberg:1987}).
Tentative reports of a steepening or cutoff of the separation distribution at
larger separations have been claimed (e.g., \citealt{Yoo:2004,Quinn:2009}), but
these claims are limited by small numbers of high-probability pairs at large
separations.

@@

Indeed, observational work that followed appears to find tentative evidence for
the existence of a second peak in the separation distribution of
radial-velocity-confirmed comoving stars (\citealt{Shaya:2011}; hereafter SO11).
A key aspect of this work was to move from searching for common proper motion
pairs to performing a statistical inference of the difference in full-space
velocity vector for pairs of stars.
At separations of $1$--$10~\pc$, star pairs in the local solar neighborhood
subtend large angles on the sky: even if the two component stars were moving
with the same 3D velocity, their proper motions and radial velocities can be
different.
By compiling radial velocities from the literature for \project{Hipparcos}
stars, SO11 applied this method to search for all comoving star pairs with
separations larger than $0.01~\pc$ within $100~\pc$ of the sun.
They find $\sim 200$ high probability comoving pairs with a separation
distribution that falls off until $\sim 1~\pc$, where the number of pairs then
begins rising in bins of log-separation.

@@

To select candidate pairs for follow-up spectroscopic observations, we only
consider mutually exclusive star pairs with Heliocentric distances $d <
200~\pc$.
With the SNR cut used above, and within the limits (e.g., magnitude range) of
the \tgas\ catalog itself, \tgas\ is $\approx$90\% complete to $200~\pc$
(\citealt{Bovy:2017}).
From this source target list, we randomly observed targets.
\figurename~\ref{fig:sample-cmd} shows a color-magnitude diagram of all targets
observed in this observing run (markers) plotted over the density of 195,294
\tgas\ stars within $200~\pc$ with \acronym{2MASS} photometry.
Candidate comoving pairs are connected by (arbitrarily colored) lines.

@@

Spectra were obtained using the \project{Modspec} spectrograph mounted on the
$2.4~{\rm m}$ Hiltner telescope at MDM
observatory\footnote{\url{http://mdm.kpno.noao.edu}} on Kitt Peak (Arizona).
\project{Modspec} was set up in long-slit mode with a 1\arcsec\ slit, a
$600~{\rm line}~{\rm mm}^{-1}$ grating, and a $2048\times2048~{\rm pixel}$ SITe
CCD detector (``Echelle'').
Only the central 300 pixel columns were read out from the CCD, and pixel rows
$>1600$ (at the blue end) were later trimmed because of the poor quantum
efficiency of the CCD at wavelengths $\lesssim 3600~{\rm \AA}$.
The resolution and wavelength range were $\approx2~{\rm \AA}~{\rm pixel}^{-1}$
from $\approx 3600$--$7200~{\rm \AA}$.
At this resolution, to obtain high signal-to-noise ($>100$ per pixel) spectra,
most exposures were between $30$--$120~{\rm s}$.
To maximize the number of observed pairs, we chose not to take
comparison lamp spectra at each pointing as it adds an overhead of $\approx
90~{\rm s}$ per observation.
All of our targets were observed at an airmass $\sec z < 1.5$.
We are primarily interested in relative velocities between pair members and
therefore don't need precise absolute calibrations of their radial velocities.
To determine rough nightly wavelength solutions, calibration spectra were
obtained using Hg-Ne and Ne comparison lamps at the beginning and end of each
night.
Additionally, atmospheric emission lines were used to correct the wavelength
solution on a per-object basis.

@@

\begin{table}[h]
    \centering
    \caption{Description of columns in data table containing separations and
    relative radial velocities for all observed groups, along with likelihood
    ratios computed with and without relative radial velocities.}
    \label{tbl:data-pairs}
    \begin{tabular}{l|l|l}
        \toprule
        Column Name    & Unit & Description\\
        \midrule
        \texttt{Oh17\_group\_id}    &      & Group ID from candidate catalog (\citealt{Oh:2017})\\
        \texttt{prob}               &      & Probability of being comoving given 3D velocity data.\\
        \texttt{sep\_tan}           & \pc  & Median of tangential separation error distribution\\
        \texttt{sep\_tan\_err}      & \pc  & Uncertainty in tangential separation\\
        \texttt{relative\_rv}       & \kms & Relative radial velocity between the two stars\\
        \texttt{relative\_rv\_err}  & \kms & Uncertainty in relative radial velocity\\
        \bottomrule
    \end{tabular}
\end{table}
